---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bubbler

<!-- badges: start -->
<!-- badges: end -->

bubbler generates amplicon sequencing bubble and barplots 
to visualize community composition. This is a common visualization in microbial ecology studies, but current implementations are hard to customize, which has prompted me to develop bubbler. This package works alongside [phyloseq](https://github.com/joey711/phyloseq) and has been designed to work with 
data generated by [dada2](https://github.com/benjjneb/dada2). Future versions are planned
to work with [qiime2](https://github.com/qiime2/qiime2) output. 

Features

- Calculate relative abundance by sample or variable.
- Filter by taxonomic level and optionally, specific samples. 
- Pool taxa below a set threshold.
- Choose how many taxa you want to display. 

Future Features

- Arrange taxa in ascending or descending order. 
- *Italicize* genus and species-level taxonomies.
- Use brewer and viridis colourschemes
- Generate global colourschemes, which are consistent between plots.
- Import multiple file types (raw tables, physeq objects, qiime2 artifacts,)
    - raw tables
    - physeq objects
    - qiime2 artifacts
    - biome files

## Installation

You can install the development version of bubbler from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("zjardyn/bubbler")
```

## Example
```{r, echo=FALSE}
load("data/physeq1.rda")
```

Lets start by making a barplot from a toy amplicon dataset. We will generate a relative abundance table, filter it at the taxonomic level of genus and subset some samples of interest (a-c). Next, we will choose how many taxa we want to display, which will set a threshold to pool our taxa by (d). We then plot it with ggplot2. To display 6 taxa, a threshold of 0.06% is set.

```{r bar}
library(bubbler)

a <- rel_abund(physeq1)
b <- choose_taxa_level(a, taxon_level = "Genus")
c <- choose_samples(b, smp_selection = c("Smp1", "Smp2", "Smp3", "Smp4", "Smp5"))

threshold <- choose_n_taxa(c, 6)
d <- pool_taxa(c, threshold)

d %>%
ggplot2::ggplot(ggplot2::aes(x = sample_id, y = rel_abund, fill = taxon)) +
    ggplot2::geom_bar(position = "fill", stat = "identity")
```
A bubbleplot is just a variation of a stacked-barplot. We will change some of the arguments to display everything from our relative abundance table. We inner_join d with our metadata to colour by Location. 


```{r bubble}
a <- rel_abund(physeq1)
# change taxon_level to Species, don't subset samples
b <- choose_taxa_level(a, taxon_level = "Species")

threshold <- choose_n_taxa(b, 6)
d <- pool_taxa(b, threshold)

d %>%
    dplyr::inner_join(., bubbler::meta_data_phy(physeq1), by = "sample_id") %>%
ggplot2::ggplot(ggplot2::aes(x = sample_id, y = taxon, size = rel_abund)) +
    ggplot2::geom_point(ggplot2::aes(colour = Location))

```

You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this.

You can also embed plots, for example:

```{r pressure, echo = FALSE}
plot(pressure)
```

In that case, don't forget to commit and push the resulting figure files, so they display on GitHub and CRAN.
