---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bubbler

<!-- badges: start -->
<!-- [![R-CMD-check](https://github.com/zjardyn/bubbler/workflows/R-CMD-check/badge.svg)](https://github.com/zjardyn/bubbler/actions) -->
[![Codecov test coverage](https://codecov.io/gh/zjardyn/bubbler/main/graph/badge.svg)](https://codecov.io/gh/zjardyn/bubbler?branch=main)
<!-- badges: end -->

Bubbler is a tidy approach to amplicon sequencing visualization. Bubbler makes relative abundance tables from amplicon-sequencing (16S/18S rRNA) datasets. Bubbler combines asv counts, taxonomic information, and optionally, metadata to create stacked barcharts, or bubble plots. Bubbler can import from [dada2](https://github.com/benjjneb/dada2), [qiime2](https://github.com/qiime2/qiime2), [phyloseq](https://joey711.github.io/phyloseq/) and [Bracken](https://github.com/jenniferlu717/Bracken) output. Relative abundance tables can be modified, with various methods implemented in bubbler.  

## Installation

You can install bubbler from [GitHub](https://github.com/) with:

``` r
if (!requireNamespace("devtools", quietly = TRUE)){
    install.packages("devtools")
}
devtools::install_github("zjardyn/bubbler")
```

## bubbler 

`bubbler` has three main steps: 

1. Import data into a relative abundance table.
2. Modify the `rel_abund` table. 
3. Make a `ggplot2` object, with stacked-bar or bubble plot aesthetics.
    
```{r, echo=FALSE, message=FALSE}
library(magrittr)
```

Here, using `tidyverse` notation, I import data from a phyloseq object, modify the table, and plot it. 

```{r, warning=FALSE, message=FALSE}
library(bubbler)

rel_abund_phy(physeq, taxa_level = "Genus") %>% # 1. import data into rel_abund.
    pool_taxa(n_taxa = 16) %>% # 2. modify  rel_abund pool to 16 most abundant taxa
    arrange_taxa() %>% # arrange plotting variables by most abundant taxa.
    arrange_sample_by_taxa() %>% # arrange samples by most abundant taxa
    bar_plot(position = "fill") # 3. plot
```


The [bubbler documentation](https://zjardyn.github.io/bubbler/) has examples for importing data, constructing rel_abund tables, and other aspects of the package. 



<!-- You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. -->



<!-- In that case, don't forget to commit and push the resulting figure files, so they display on GitHub and CRAN. -->
