---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bubbler

<!-- badges: start -->
<!-- badges: end -->

bubbler generates amplicon sequencing bubble and barplots 
to visualize community composition. This is a common visualization in microbial ecology studies, but current implementations are hard to deeply customize, which has prompted me to develop bubbler. This package works alongside [phyloseq](https://github.com/joey711/phyloseq) and has been designed to work with 
data generated by [dada2](https://github.com/benjjneb/dada2). Future versions are planned
to work with [qiime2](https://github.com/qiime2/qiime2) output. 

Features

- Calculate relative abundance by sample or variable.
- Filter by taxonomic level and optionally, specific samples. 
- Pool taxa below a set threshold.
- Choose how many taxa you want to display. 

Future Features

- Arrange samples and taxa.
- *Italicize* genus and species-level taxonomies.
- Use brewer and viridis colourschemes
- Generate global colourschemes, which are consistent between plots.
- Import multiple file types:
    - raw tables
    - physeq objects
    - qiime2 artifacts
    - biome files

## Installation

You can install the development version of bubbler from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("zjardyn/bubbler")
```

You will also need phyloseq, which can be installed with:

```r
if(!requireNamespace("BiocManager")){
  install.packages("BiocManager")
}
BiocManager::install("phyloseq")
```

## Sample-wise plots  
```{r, echo=FALSE}
load("data/physeq1.rda")
```

```{r barfill, message=FALSE}
library(bubbler)
library(phyloseq)
library(tidyverse)

a <- rel_abund(phy = physeq1, taxa_level = "Genus", meta_data = FALSE)
sample_selection <- c("Smp1", "Smp2", "Smp3", "Smp4", "Smp5")
b <- choose_samples_rel_abund(a, sample_selection)
 
threshold <- choose_n_taxa(b, 6)
c <- pool_taxa(b, threshold)

ggplot(c, aes(x = sample_id, y = rel_abund, fill = taxon)) +
    geom_bar(position = "fill", stat = "identity")
```

```{r barstack}
ggplot(c, aes(x = sample_id, y = rel_abund, fill = taxon)) +
    geom_bar(position = "stack", stat = "identity")
```

```{r barfill_var}
a <- rel_abund(phy = physeq1, taxa_level = "Genus", meta_data = TRUE)
threshold <- choose_n_taxa(a, 6)
b <- pool_taxa(a, threshold)

metadata <- meta_data_phy(physeq1)
c <- inner_join(b, metadata, by = "sample_id")
    
ggplot(c, aes(x = Location, y = rel_abund, fill = taxon)) +
    geom_bar(position = "fill", stat = "identity")
```

```{r barstack_var}
ggplot(c, aes(x = Location, y = rel_abund, fill = taxon)) +
    geom_bar(position = "stack", stat = "identity")
```





<!-- You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. -->



<!-- In that case, don't forget to commit and push the resulting figure files, so they display on GitHub and CRAN. -->
