---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bubbler

<!-- badges: start -->
<!-- badges: end -->

bubbler generates amplicon sequencing bubble and barplots 
to visualize community composition. This is a common visualization in microbial ecology studies, but current implementations are hard to deeply customize, which has prompted me to develop bubbler. This package works alongside [phyloseq](https://github.com/joey711/phyloseq) and has been designed to work with 
data generated by [dada2](https://github.com/benjjneb/dada2). Future versions are planned
to work with [qiime2](https://github.com/qiime2/qiime2) output. 

Features

- Calculate relative abundance by sample or variable.
- Filter by taxonomic level and optionally, specific samples. 
- Pool taxa below a set threshold.
- Choose how many taxa you want to display. 

Future Features

- Arrange samples and taxa.
- *Italicize* genus and species-level taxonomies.
- Use brewer and viridis colourschemes
- Generate global colourschemes, which are consistent between plots.
- Import multiple file types:
    - raw tables
    - physeq objects
    - qiime2 artifacts
    - biome files

## Installation

You can install the development version of bubbler from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("zjardyn/bubbler")
```

You will also need phyloseq, which can be installed with:

```r
if(!requireNamespace("BiocManager")){
  install.packages("BiocManager")
}
BiocManager::install("phyloseq")
```

## Sample-wise plots  
```{r, echo=FALSE}
load("data/physeq1.rda")
```


Lets start by making a barplot from a toy amplicon dataset. We will generate a relative abundance table, filter it at the taxonomic level of genus and subset some samples of interest (a-c). Next, we will choose how many taxa we want to display, which will set a threshold to pool our taxa by (d). We then plot it with ggplot2. To display 6 taxa, a threshold of 0.06% is set.


```{r barfill}
library(bubbler)
library(ggplot2)

a <- rel_abund(physeq1)
b <- choose_taxa_level(a, taxon_level = "Genus")
c <- choose_samples(b, smp_selection = c("Smp1", "Smp2", "Smp3", "Smp4", "Smp5"))

threshold <- choose_n_taxa(c, 6)
d <- pool_taxa(c, threshold)

d %>%
ggplot(aes(x = sample_id, y = rel_abund, fill = taxon)) +
    geom_bar(position = "fill", stat = "identity")
```


The geom_bar uses the position argument of fill, which scales our bars to fill the y axis of rel_abund. While this looks good, we are missing information on relative abundances between samples. We can use position = "stack" to see this information.  


```{r barstack}
d %>%
ggplot(aes(x = sample_id, y = rel_abund, fill = taxon)) +
    geom_bar(position = "stack", stat = "identity")
```


This clears up a misconception about setting our threshold. The first plot is set to 0.06% but the pooled bar appears larger than that on the y axis. The second plot shows the true relative abundance between samples, where our pooled taxon is smaller.  


To pool, internally we are looking at the max value for a given taxon and, if that value is less than 0.06, it is pooled. 


A bubbleplot is just a variation of a stacked-barplot. We will change some of the arguments to display everything from our relative abundance table. We inner_join d with some metadata to colour by Location. 


```{r bubble}
a <- rel_abund(physeq1)
# change taxon_level to Species, don't subset samples
b <- choose_taxa_level(a, taxon_level = "Species")

threshold <- choose_n_taxa(b, 12)
d <- pool_taxa(b, threshold)

d %>%
    dplyr::inner_join(., bubbler::meta_data_phy(physeq1), by = "sample_id") %>%
ggplot2::ggplot(ggplot2::aes(x = sample_id, y = taxon, size = rel_abund)) +
    ggplot2::geom_point(ggplot2::aes(colour = Location))

```


## Variable-wise plots


If you have a variable of interest from your metadata which you would rather use as the x-axis, you can use rel_abund_var to specify it. You must also specify it in pool_taxa.

```{r}
a <- rel_abund_var(physeq1, var = "Location")
b <- choose_taxa_level(a, taxon_level = "Genus")
threshold <- choose_n_taxa(b, 6)
c <- pool_taxa(b, threshold, var = "Location")
c %>%
ggplot2::ggplot(ggplot2::aes(x = Location , y = rel_abund, fill = taxon)) +
    ggplot2::geom_bar(position = "fill", stat = "identity")
```





You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this.

You can also embed plots, for example:

```{r pressure, echo = FALSE}
plot(pressure)
```

In that case, don't forget to commit and push the resulting figure files, so they display on GitHub and CRAN.
