---
title: "2. Calculate relative abundance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{calc_rel_abund}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  fig.width = 7
  
)
```

```{r setup, echo=FALSE}
library(bubbler)
```
# Relative abundance

`bubbler` takes an ASV table and converts counts to relative abundances using this simple formula:

$\text{Relative abundance} = \frac{x_i}{\sum x_i} = 1$

Where $x_i$ is the count of a given ASV, divided by the sum of all ASV counts. This way, the sum of all relative abundances equal one.   \n   `bubbler` then tacks on taxonomic information and optionally metadata. By default, `bubbler::bar_plot` produces stacked bar plots, which show the proportional differences of counts between samples. 


```{r load_qiime, echo = FALSE, message=FALSE, warning=FALSE }
library(ggplot2)
library(viridis)
# path to qiime-formatted asv counts 
counts_q <- system.file("extdata", "qiime", "table-dada2.qza", package = "bubbler")

# path to qiime-formatted taxonomy data 
taxa_q <- system.file("extdata", "qiime", "taxonomy.qza", package = "bubbler")

# path to qiime-formatted metadata
metadata_q <- system.file("extdata", "qiime", "sample-metadata.tsv", package = "bubbler")

```

```{r fig.height= 6 }
# 1. make rel_abund
rel_abund  <-  rel_abund_qiime(counts_q, 
                               taxa_q, 
                               metadata_q,
                               taxa_level = "Genus")

# 2. modify rel_abund
rel_abund_pool <- rel_abund %>%
    # pool taxa so that only the 12 most abundant taxa are displayed 
    pool_taxa(n_taxa = 12, keep_metadata = TRUE) 

# 3. plot rel_abund
rel_abund_pool %>%
    bar_plot()  + 
    # facet samples by body site. free_x removes unwanted white space
    facet_wrap(~body_site, scales = "free_x" ) 

```



`bubbler::bar_plot`s can also be filled in. Based on the above plot, subject proportions on the right palm are vastly different, and this information would be lost in a filled-in bar plot. By filling the plotting area, we get a better representation of the between-sample composition, at the cost of obfuscating the between-sample proportions. 

```{r}
rel_abund_pool %>%
    bar_plot(position = "fill")  + 
    facet_wrap(~body_site, scales = "free_x") 
```



One last thing about relative abundances: when setting `position` to "fill", we are effectively recomputing rel_abund, so that it sums to one within each bar on the x-axis. This transformation is equivalent to the following formula:

$\text{Grouped relative abundance} = \frac{x_{ij}}{\sum_{j=1}^{k} x_{ij}} = k \text{ levels}$

Where $k$ is the levels of the grouping variable (in this case `sample_id`) and $_i$ indexs through the ASVs and $_j$ indexs through the levels, Within each level of the grouping variable, the sum of relative abundance equals one, and the total sum of rel_abund is equal to the $k$ levels. This information is not very important in practice because `position` = "fill" scales the relative abundances for you.

## Visualizing variables 

`bubbler`  allows you to focus on any metadata variable in your relative abundance table. Say you wanted to look at relative abundance across `body_site`, rather than `sample_id` on the x-axis. This can be done by modifying the `x_var` argument of `bar_plot`.


```{r}
rel_abund_qiime(counts_q, taxa_q, metadata_q, taxa_level = "Genus") %>%
    pool_taxa(n_taxa = 12,
              keep_metadata = TRUE) %>%
    arrange_taxa() %>%
    bar_plot(x_var = "body_site")
```



For any given `x_var`, (`sample_id`, `body_site`, etc ), you have two options for computing the scaled rel_abund: Setting the `var` argument within the `rel_abund` function, or setting `position` = "fill" within `bubbler::bar_plot`. I prefer the latter method, which keeps the proportional differences.


```{r, eval=FALSE}
# compute in rel_abund
rel_abund_qiime(counts_q, taxa_q, metadata_q,
                taxa_level = "Genus",
                var = "body_site") %>% #  scaling by body_site
    pool_taxa(n_taxa = 12,
              keep_metadata = TRUE) %>%
    arrange_taxa() %>%
    bar_plot(x_var = "body_site") # setting body_site as x_var
```

```{r}
# scale in plot
rel_abund_qiime(counts_q, taxa_q, metadata_q,
                taxa_level = "Genus") %>%
    pool_taxa(n_taxa = 12,
              keep_metadata = TRUE) %>%
    arrange_taxa() %>%
    bar_plot(position = "fill", x_var = "body_site")  # setting position as fill
```

## Showing both scaled and unscaled information.

If you want the best of both worlds, you can create a scaled bar plot, and superimpose the proportional abundances. Currently, this feature does not work with a faceting functions, so the only option would be to split a rel_abund table into multiple plots.

```{r, message=FALSE, warning=FALSE, fig.width= 9, fig.width= 9}
# order by sample read abundance, plot as line
library(patchwork)

counts_q <- system.file("extdata", "qiime", "table-dada2.qza", package = "bubbler")
taxa_q <- system.file("extdata", "qiime", "taxonomy.qza", package = "bubbler")
metadata_q <- system.file("extdata", "qiime", "sample-metadata.tsv", package = "bubbler")

q <- rel_abund_qiime(
    asv_qiime = counts_q,
    taxa_qiime = taxa_q,
    metadata_qiime = metadata_q,
    taxa_level = "Genus", ) %>% 
    pool_taxa(n_taxa = 8, keep_metadata = TRUE) %>%
    arrange_taxa() %>%
    arrange_var(levels = "body_site")
    # arrange_var_abund(flip = TRUE)

yes <- subset_rel_abund(q, var = "reported_antibiotic_usage", selection = "Yes")
no <- subset_rel_abund(q, var = "reported_antibiotic_usage", selection = "No")

p1 <- bar_plot(yes, position = "fill", true_line = TRUE ) + ggtitle("Antibiotics used") + 
geom_text(aes(label = body_site, y = 0.9), 
          angle = 90, color = "white", size = 3)
p2 <- bar_plot(no, position = "fill" , true_line = TRUE) + ggtitle("Antibiotics not used") + 
geom_text(aes(label = body_site, y = 0.9), 
          angle = 90, color = "white", size = 3)
 p1 + p2 + plot_layout(guides = "collect", axes = "collect")

```



