---
title: "1. Data Import"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{data_import}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Amplicon sequencing input can be variable so `bubbler` has many data import options. 

```{r setup}
library(bubbler)
```

## Qiime2

[qiime2](https://qiime2.org/) asv-table and taxonomy artifacts (.qza), and optionally, the qiime-formatted metadata (.tsv) can be imported. This example uses data from the "Moving Pictures" Qiime2 [tutorial](https://docs.qiime2.org/2024.5/tutorials/moving-pictures/).

```{r}
# path to qiime-formatted asv counts 
counts_q <- system.file("extdata", "qiime", "table-dada2.qza", package = "bubbler")
# path to qiime-formatted taxonomy data 
taxa_q <- system.file("extdata", "qiime", "taxonomy.qza", package = "bubbler")
# path to qiime-formatted metadata
metadata_q <- system.file("extdata", "qiime", "sample-metadata.tsv", package = "bubbler")

# make a relative abundance table 
rel_abund_qiime(counts_q, taxa_q, metadata_q)
```

## DADA2/TSV

DADA2 outputs can be an asv-table and taxonomic classifications as .tsv files. An asv-table has ASVs as columns (variables) and samples as rows (observations), with the cells being counts. A taxonomy table has taxonomic levels as columns and ASVs as rows. 
```{r}
# path to asv counts in tab-separated format
counts <- system.file("extdata", "tsv", "seqtab.tsv", package = "bubbler")
# path to taxonomy data in tab-separated format
taxa <- system.file("extdata", "tsv", "taxa.tsv", package = "bubbler")
# path to metadata in tab-seperated format 
metadata <- system.file("extdata", "tsv", "metadata.tsv", package = "bubbler")

# make a relative abundance table
rel_abund_tsv(counts, taxa, metadata)
```

## Phyloseq

If your data is in the form of a `phyloseq` object, you can import it. `bubbler` expects to see a otu_table and tax_table, and optionally, sam_data.

```{r}
# example phyloseq
rel_abund_phy(physeq, taxa_data = TRUE, meta_data = TRUE)
```


## Kracken2/Bracken 

The .txt files from a [Bracken](https://github.com/jenniferlu717/Bracken) workflow can be imported.


```{r}
# path to bracken-formatted .txt files
path <- system.file("extdata", "bracken", package = "bubbler")

rel_abund_bracken(path)
```

