---
title: "1. Data Import"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{data_import}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Data import can be a frustrating first step. bubbler tries to alleviate this by giving the user flexibility with how they can get their data into R. 

Load bubbler and import some data. 

```{r setup}
library(bubbler)
```

## Qiime2

If you have just ran your amplicon sequencing data through [qiime2](https://qiime2.org/), you can import the asv-table and taxonomy artifacts (.qza), and optionally, the qiime-formatted metadata (.tsv). Here, I show importing, using data from the "Moving Pictures" Qiime2 [tutorial](https://docs.qiime2.org/2024.5/tutorials/moving-pictures/).

```{r}
# path to qiime-formatted asv counts 
counts_q <- system.file("extdata", "qiime", "table-dada2.qza", package = "bubbler")

# path to qiime-formatted taxonomy data 
taxa_q <- system.file("extdata", "qiime", "taxonomy.qza", package = "bubbler")

# path to qiime-formatted metadata
metadata_q <- system.file("extdata", "qiime", "sample-metadata.tsv", package = "bubbler")

# make a relative abundance table 
rel_abund_qiime(counts_q, taxa_q, metadata_q)

```

## DADA2 

DADA2 allows you to generate an asv-table and taxonomic classifications from amplicon sequencing data, all within R. An asv-table has ASVs as columns (variables) and samples as rows (observations). A taxonomy table has taxonomic levels as columns and ASVs as rows. `bubbler` can import these files in the form of .tsv files, with optional metadata as a .tsv as well. 

```{r}
# path to asv counts in tab-separated format
counts <- system.file("extdata", "tsv", "seqtab.tsv", package = "bubbler")

# path to taxonomy data in tab-separated format
taxa <- system.file("extdata", "tsv", "taxa.tsv", package = "bubbler")

# path to metadata in tab-seperated format 
metadata <- system.file("extdata", "tsv", "metadata.tsv", package = "bubbler")

# make a relative abundance table
rel_abund_tsv(counts, taxa, metadata)
```

## Phyloseq

If you can get your data into a phyloseq object, it is possible to import that as well. `bubbler` expects to see a otu_table and tax_table, and optionally, sam_data within the phyloseq object. 

```{r}
# example phyloseq
rel_abund_phy(physeq1,
              taxa_data = TRUE,
              meta_data = TRUE)
```





