---
title: "data_import"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{data_import}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Data import can be a frustrating first step. `bubbler` tries to alleviate this frustration by giving the user flexibility with how they choose to get their data into R. Let's load `bubbler` and import some data! 

```{r setup}
library(bubbler)
```

## Qiime2

If you have just ran your amplicon sequencing data through [qiime2](https://qiime2.org/), you can import the asv-table and taxonomy artifacts (.qza), and optionally, the qiime-formatted metadata (.tsv). If you are unfamiliar with qiime2, check out their documentation, specifically the [Moving Pictures](https://docs.qiime2.org/2024.5/tutorials/moving-pictures/) tutorial. Let's load the data from that tutorial!

```{r}
# path to qiime-formatted asv counts 
counts_q <- system.file("extdata", "qiime", "table-dada2.qza", package = "bubbler")

# path to qiime-formatted taxonomy data 
taxa_q <- system.file("extdata", "qiime", "taxonomy.qza", package = "bubbler")

# path to qiime-formatted metadata
metadata_q <- system.file("extdata", "qiime", "sample-metadata.tsv", package = "bubbler")

# make a relative abundance table 
rel_abund_qiime(counts_q, taxa_q, metadata_q)

```

## DADA2 or 




