---
title: "3. Pooling and arranging" 
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pool_taxa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Pooling 

```{r setup, message=FALSE, warning=FALSE}
library(bubbler)
library(dplyr)
```

```{r load_qiime, echo=FALSE}
# path to qiime-formatted asv counts 
counts_q <- system.file("extdata", "qiime", "table-dada2.qza", package = "bubbler")

# path to qiime-formatted taxonomy data 
taxa_q <- system.file("extdata", "qiime", "taxonomy.qza", package = "bubbler")

# path to qiime-formatted metadata
metadata_q <- system.file("extdata", "qiime", "sample-metadata.tsv", package = "bubbler")
```

The taxonomic level can be selected with the `taxa_level` argument, when generating 
the relative abundance table. 

```{r}
rel_abund <- rel_abund_qiime(counts_q, taxa_q, metadata_q, taxa_level = "Genus")

rel_abund %>%
    summarise(unique_taxa = n_distinct(taxon)) # see how many unique taxa there are 
```

There are 196 unique taxa in the relative abundance table. If we attempt to plot this 
right away, the legend will include all these taxa. The table must be filtered 
to pool taxa below a given threshold. The `pool_taxa` function can be given a numerical
threshold or the number of taxa that should be displayed. In the latter case, the
threshold is determined automatically. 

```{r}
rel_abund %>%
    pool_taxa(n_taxa = 12) %>% # threshold set to 0.0038
    bar_plot(position = "fill")
```


Pooling, by default will remove any meta data that was attached to the relative abundance table. Specify `keep_metadata` to keep it.

```{r}
rel_abund %>% 
    pool_taxa(n_taxa = 16,
              keep_metadata = TRUE) %>%
    bar_plot(position = "fill", x_var = "body_site")
    
```

# Arranging 


