---
title: "3. Pooling and arranging" 
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pool_taxa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Pooling taxonomy

```{r setup, message=FALSE, warning=FALSE}
library(bubbler)
library(ggplot2)
```

```{r load_qiime, echo=FALSE}
# path to qiime-formatted asv counts 
counts_q <- system.file("extdata", "qiime", "table-dada2.qza", package = "bubbler")

# path to qiime-formatted taxonomy data 
taxa_q <- system.file("extdata", "qiime", "taxonomy.qza", package = "bubbler")

# path to qiime-formatted metadata
metadata_q <- system.file("extdata", "qiime", "sample-metadata.tsv", package = "bubbler")
```

The taxonomic level is selected with the `taxa_level` argument, when generating 
the relative abundance table. By default it is set to "Phylum". Let's see how many different taxa are present in a table. 


```{r, message=FALSE}
library(dplyr)
rel_abund <- rel_abund_qiime(counts_q, taxa_q, metadata_q, taxa_level = "Genus")

rel_abund %>%
    summarise(unique_taxa = n_distinct(taxon)) # see how many unique taxa there are 
```

There are 196 unique taxa in the relative abundance table. If we attempt to plot this 
right away, the legend will include all these taxa. The table must be filtered 
to pool taxa below a given threshold. The `pool_taxa` function can be given a numerical
threshold or the number of taxa that should be displayed. In the latter case, the
threshold is determined automatically. 

```{r}
rel_abund %>%
    pool_taxa(n_taxa = 12) %>% # threshold set to 0.0038
    bar_plot(position = "fill") 
```


Pooling, by default will remove any meta data that was attached to the relative abundance table. Specify `keep_metadata` to keep it. Setting `label` to FALSE will set the threshold label in the legend to "Other". 

```{r}
rel_abund %>% 
    pool_taxa(n_taxa = 8,
              keep_metadata = TRUE,
              label = FALSE) %>%
    bar_plot(position = "fill", x_var = "body_site")
```


# Arranging 


## Taxa

In the legend, taxa are ordered alphabetically, which is the default for `ggplot2` when dealing with character variables.  Sorting the taxa by their abundance would be more meaningful. Using the `arrange_taxa` function, taxa can be arranged in ascending order, or flipped by setting  the `order` argument to "top" or "bottom". The threshold order can also be changed by setting the `pooled` argument to "top" or "bottom". After ordering, it becomes clear that *Bacteroides* is the most abundant taxon in the gut, and *Streptococcus* among the other samples.

```{r}
rel_abund %>% 
    pool_taxa(n_taxa = 8,
              keep_metadata = TRUE) %>%
    arrange_taxa(pooled = "top", order = "bottom") %>% # arrange by abundance
    bar_plot(position = "fill", x_var = "body_site") +  
    labs(y = "mean_rel_abund")
```


## Variable

Another option is to arrange the samples by a variable, which can be useful when the data includes metadata. Observe the similarities between the above and below plots. The above plot represents the mean relative abundances among the levels of `body_site`, and the bottom plot represents the relative abundance by sample, ordered by `body_site`. 


```{r}
rel_abund %>% 
    pool_taxa(n_taxa = 8,
              keep_metadata = TRUE) %>%
    arrange_taxa(pooled = "top", order = "bottom") %>% # arrange by abundance
    arrange_var(var = "sample_id", levels = "body_site") %>%
    bar_plot(position = "fill") 
```


## Samples by taxa abundance

Samples can be arranged by taxa abundance. I first saw this technique used by Chenxin Li, in their github repo: [Friends Dont Let Friends Make Bad Graphs](https://github.com/cxli233/FriendsDontLetFriends/tree/main#13-friends-dont-let-friends-forget-to-reorder-stacked-bar-plot). Combined with `bubbler::arrange_taxa`, `bubbler::arrange_sample_by_taxa` will partition samples by their most abundant taxa. In the absence of meta data, this is a good preliminary visualization to produce. 

```{r}
rel_abund %>% 
    pool_taxa(n_taxa = 16) %>%
    arrange_taxa() %>% 
    arrange_sample_by_taxa() %>%
    bar_plot(position = "fill") 
```

<!-- TODO: variable by abundance  -->
