---
title: "3. Pooling and arranging" 
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pool_taxa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Pooling 

```{r setup, message=FALSE, warning=FALSE}
library(bubbler)
library(dplyr)
library(ggplot2)
```

```{r load_qiime, echo=FALSE}
# path to qiime-formatted asv counts 
counts_q <- system.file("extdata", "qiime", "table-dada2.qza", package = "bubbler")

# path to qiime-formatted taxonomy data 
taxa_q <- system.file("extdata", "qiime", "taxonomy.qza", package = "bubbler")

# path to qiime-formatted metadata
metadata_q <- system.file("extdata", "qiime", "sample-metadata.tsv", package = "bubbler")
```

The taxonomic level can be selected with the `taxa_level` argument, when generating 
the relative abundance table. 

```{r}
rel_abund <- rel_abund_qiime(counts_q, taxa_q, metadata_q, taxa_level = "Genus")

rel_abund %>%
    summarise(unique_taxa = n_distinct(taxon)) # see how many unique taxa there are 
```

There are 196 unique taxa in the relative abundance table. If we attempt to plot this 
right away, the legend will include all these taxa. The table must be filtered 
to pool taxa below a given threshold. The `pool_taxa` function can be given a numerical
threshold or the number of taxa that should be displayed. In the latter case, the
threshold is determined automatically. 

```{r}
rel_abund %>%
    pool_taxa(n_taxa = 12) %>% # threshold set to 0.0038
    bar_plot(position = "fill") + 
    scale_fill_viridis_d(option = "turbo")
```


Pooling, by default will remove any meta data that was attached to the relative abundance table. Specify `keep_metadata` to keep it.

```{r}
rel_abund %>% 
    pool_taxa(n_taxa = 16,
              keep_metadata = TRUE) %>%
    bar_plot(position = "fill", x_var = "body_site") + 
   scale_fill_viridis_d(option = "turbo")
    
```


# Arranging 


## Taxa

By default taxa are in alphabetical order, which can be seen in the legend of the plot.  Sorting the taxa by their abundance is a more meaningful sorting strategy. By default, taxa are arranged in ascending order but this can be switched with the `order` argument. The threshold position can also be modified with the `pooled` argument. With this ordering, it becomes clear that *Bacteroides* is the most abundant taxon in the gut, and *Streptococcus* among the other samples.

```{r}
rel_abund %>% 
    pool_taxa(n_taxa = 8,
              keep_metadata = TRUE) %>%
    arrange_taxa(pooled = "top", order = "bottom") %>% # arrange by abundance
    bar_plot(position = "fill", x_var = "body_site") +  
       scale_fill_viridis_d(option = "turbo")
```


## Variable

Another option is to arrange the samples by a variable, which can be useful when the data includes metadata. Observe the similarities between the above and below plots. 


```{r}
rel_abund %>% 
    pool_taxa(n_taxa = 8,
              keep_metadata = TRUE) %>%
    arrange_taxa(pooled = "top", order = "bottom") %>% # arrange by abundance
    arrange_var(var = "sample_id", levels = "body_site") %>%
    bar_plot(position = "fill") + 
        scale_fill_viridis_d(option = "turbo")
```


## Samples by abundance

Samples can be arranged by taxa abundance. This is a good preliminary plot to generate to visualize the abundance of taxa within all the samples. 

```{r}
rel_abund %>% 
    pool_taxa(n_taxa = 16) %>%
    arrange_taxa() %>% 
    arrange_sample_by_taxa() %>%
    bar_plot(position = "fill") + 
    scale_fill_viridis_d(option = "turbo")
```

<!-- TODO: variable by abundance  -->
